General limits

### 2017-09-08 papers/talk-7/paper-7.tex
reworked construction of arbitrary limits

\begin{lemma}
  \label{finiteproducts}
  Let $\mathscr C$ be a category with a terminal object $1$ and binary products.
  Then we can construct all finite products.
\end{lemma}
\begin{proof}
  \begin{itemize}
  \item The empty product is just the terminal object by
    \hyperref[emptyproduct]{Example \ref*{emptyproduct}}
  \item The unary product of any object $A$ is constructed by the binary product
    $A\times 1$ %% TODO: explain why this works...
  \item The binary product exists by assumption
  \item We want to build some finite product over objects $D_1, \dots D_k$.
    Assume by induction that the following two products exist.
    \[ P_1 \coloneqq \displaystyle \prod_{i \in \{1, \dots, m \} } D_i \text{ , and } P_2 \coloneqq \displaystyle \prod_{i \in \{m+1, \dots, k\} } D_i \]
    with projection maps $(\pi_i)_{i \in \{1, \dots, m \} }$ for $P_1$, and $(\tau_i)_{i \in \{m+1, \dots, k\} }$ for $P_2$.
    Then there exists a binary product of $P_1 \times P_2$ with projection maps $f_{P_1}, f_{P_2}$.
    $P_1 \times P_2$ has maps $\pi_i \circ f_{P_1}$ and $\tau_i \circ f_{P_2}$.
    $P_1 \times P_2$ is therefore a cone.
    Since $P_1$ and $P_2$ are products, any other cone $V$ over them has unique maps $h_{P_1}$ and $h_{P_2}$.
    The unique morphism $V \rightarrow P_1 \times P_2$ is given by $\langle h_{P_1}, h_{P_2} \rangle$.
    $P_1 \times P_2$ is therefore universal.
  \end{itemize}
  Intuitively the construction works by stacking binary products on top of each other.
  Since we assume $\mathscr C$ to have all products, the product of two products exists. This way we can create
  arbitrarily large finite products.
\end{proof}

@@

\begin{fact}
  \label{pullbackproduct}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct finite products.
\end{fact}
\begin{proof}
Finite products can be constructed
with the terminal object and binary products by
\hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
Therefore we only show how to construct binary products.
Consider following the diagram:
\[
  \begin{tikzcd}
    & B \arrow{d} \\
    A \arrow{r} & 1
  \end{tikzcd}
\]
The pullback $U$ exists by assumption.
\[
  \begin{tikzcd}
    U \arrow{r}{p_B} \arrow{d}{p_A} & B \arrow{d}{f} \\
    A \arrow{r}{g} & 1
  \end{tikzcd}
\]
The arrows $f$ and $g$ are unique and therefore monic.
Hence $U$ satisfies the universality condition and is indeed the product.
%% TODO: possibly show this for one calculation
\end{proof}

@@

\begin{theorem}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct equalizers.
\end{theorem}
\begin{proof}
  We start with the initial diagram for equalizers.
  \[
    \begin{tikzcd}
      A \arrow[shift left]{r}{f} \arrow[shift right, swap]{r}{g} & B
    \end{tikzcd}
  \]
  We transform this into the following diagram.
  %%TODO: explain why they're equivalent
  \[
    \begin{tikzcd}
      & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  The product $B \times B$ exists by \hyperref[pullbackproduct]{Theorem \ref*{pullbackproduct}}.
  We then have the pullback $E$, such that the following diagram commutes.
  \[
    \begin{tikzcd}
      E \arrow{r}{h} \arrow{d}{e} & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  Since $E$ is a pullback the square above commutes and
  we get $f \circ e = 1_B \circ h = h = 1_B \circ h = g \circ e$, and $E$ is therefore
  indeed an equalizer of $f$ and $g$.
\end{proof}

@@

\begin{fact}
  Let $\mathscr C$ be a category that has all products and equalizers.
  Let $D : \bm I \to \mathscr C$ be any diagram.
  Then we can construct a limit of $D$.
\end{fact}
\begin{proof}
We use the following construction:
\[
  \begin{tikzcd}
    E \arrow{r}{e} &
    \displaystyle \prod_{I \in \bm{I}} D(I)
    \arrow[r,shift left, "s"] \arrow[r,shift right,swap,"t"] &
    \displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D(J)
  \end{tikzcd}
\]
with the maps $s, t$ that have components $s_n, t_n$ for all $K \overset{n}{\to} J \in \bm{I}$:
\begin{align*}
  s_n &= Dn \circ p_K\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_K}{\rightarrow} D_K \overset{Dn}{\rightarrow} D_J \\
  t_n &= p_J\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_J}{\rightarrow} D_J
\end{align*}
All products exist by assumption.
To clarify, the product over all maps $K \overset{n}{\to} J \in \bm{I}$
has components $p_n = J$ for each map $n$.\\
We take $E$ as the equalizer of $s$ and $t$ which also exists by assumption.
We therefore have $s \circ e = t \circ e$.
The maps $s$ and $t$ have components. Therefore the equality has to hold for all
components and we get
\begin{align*}
  s \circ e &= t \circ e\\
  s_n \circ e &= t_n \circ e, \quad \forall\ K \overset{n}{\rightarrow} J \in \bm I \\
  Dn \circ p_K \circ e &= p_J \circ e
\end{align*}
The goal is to construct a limit.
We do this by constructing a cone and showing that it is universal.
For the construction of a cone
we need a vertex and projection maps into every object of the diagram,
such that the commuting condition of cones is satisfied.
We take $E$ as the vertex and
$\uprho_I \coloneqq p_I \circ e, \forall\ I \in \bm I$
as the projection maps of the cone.
Clearly, we have all necessary projection maps and we only need to verify
the commuting condition of cones.
We have all maps $K \overset{n}{\rightarrow} J \in \bm I$
\begin{align*}
  Dn \circ p_K \circ e &= p_J \circ e \\
  Dn \circ \uprho_K &= \uprho_J
\end{align*}
which corresponds to the commuting diagram
\[
  \begin{tikzcd}
    E \arrow{r}{\uprho_K} \arrow{dr}[swap]{\uprho_J} & D_K \arrow{d}{Dn}\\
    & D_J
  \end{tikzcd}
\]
This is exactly the commuting condition we need for a cone.\\
Since any other cone over $D$
is also a fork of $s$ and $t$
it follows that the equalizer $E$ is also a universal cone
and therefore a limit of $D$.
\end{proof}
\begin{remark}
  We can even construct limits of infinite diagrams
  because we assumed that $\mathscr C$ has all products.
  If $\mathscr C$ only has finite products, then we can
  use this construction for
  limits of finite diagrams.\\
  Using \hyperref[equiinstance]{Section \ref*{equiinstance}}
  we can alternatively require pullbacks and a terminal object instead of finite products.
\end{remark}

>>>>

\begin{lemma}
  \label{finiteproducts}
  Let $\mathscr C$ be a category with a terminal object $1$ and binary products.
  Then we can construct all finite products.
\end{lemma}
\begin{proof}
  \begin{itemize}
  \item The empty product is just the terminal object by
    \hyperref[emptyproduct]{Example \ref*{emptyproduct}}
  \item The unary product of any object $A$ is constructed by the binary product
    $A \times A$.
    We have $1_A \from A \to A$, and $\pi_1, \pi_2 \from A \times A \to A$.
    By the universal property of products there exists a unique map $h \from A \to A \times A$.
    We have $1_A = \pi_1 \circ h$.\\
    On the other hand we have $(\pi_1 \circ h) \circ \pi_1 = 1_a \circ \pi_1 = \pi_1 \circ ( h \circ \pi_1)$,
    and therefore $1_A = h \circ \pi_1$.
    We have constructed an isomorphism and therefore $A \times A \iso A$.
  \item The binary product exists by assumption for all objects $A, B$.
  \item We want to build some finite product over objects $D_1, \dots D_k$.
    Assume by induction that the following two products exist.
    \[ P_1 \coloneqq \displaystyle \prod_{i \in \{1, \dots, m \} } D_i \text{ , and } P_2 \coloneqq \displaystyle \prod_{i \in \{m+1, \dots, k\} } D_i \]
    with projection maps $(\pi_i)_{i \in \{1, \dots, m \} }$ for $P_1$, and $(\tau_i)_{i \in \{m+1, \dots, k\} }$ for $P_2$.
    Then there exists a binary product of $P_1 \times P_2$ with projection maps $f_{P_1}, f_{P_2}$.
    $P_1 \times P_2$ has maps $\pi_i \circ f_{P_1}$ and $\tau_i \circ f_{P_2}$.
    $P_1 \times P_2$ is therefore a cone.
    Since $P_1$ and $P_2$ are products, any other cone $V$ over them has unique maps $h_{P_1}$ and $h_{P_2}$.
    The unique morphism $V \rightarrow P_1 \times P_2$ is given by $\langle h_{P_1}, h_{P_2} \rangle$.
    $P_1 \times P_2$ is therefore universal.
  \end{itemize}
  Intuitively the construction works by stacking binary products on top of each other.
  Since we assume $\mathscr C$ to have all products, the product of two products exists. This way we can create
  arbitrarily large finite products.
\end{proof}

@@

\begin{fact}
  \label{pullbackproduct}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct finite products.
\end{fact}
\begin{proof}
Finite products can be constructed
with the terminal object and binary products by
\hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
Therefore we only show how to construct binary products.
Consider following the diagram:
\[
  \begin{tikzcd}
    & B \arrow{d} \\
    A \arrow{r} & 1
  \end{tikzcd}
\]
The pullback $U$ exists by assumption.
\[
  \begin{tikzcd}
    U \arrow{r}{p_B} \arrow{d}{p_A} & B \arrow{d}{f} \\
    A \arrow{r}{g} & 1
  \end{tikzcd}
\]
Since $U$ is a pullback we have for any other cone $V$ with projections $q_A, q_B$
a unique map $h \from V \to U$, such that
$q_A = p_A \circ h$, and $q_B = p_B \circ h$, which is exactly the commuting condition
for products.
\end{proof}

@@

\begin{theorem}
  \label{pullbackequalizer}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct equalizers.
\end{theorem}
\begin{proof}
  We start with the initial diagram for equalizers.
  \[
    \begin{tikzcd}
      A \arrow[shift left]{r}{f} \arrow[shift right, swap]{r}{g} & B
    \end{tikzcd}
  \]
  We transform this into the following diagram.
  \[
    \begin{tikzcd}
      & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  The product $B \times B$ exists by \hyperref[pullbackproduct]{Theorem \ref*{pullbackproduct}}.
  Furthermore we have $B \times B \iso B$ by \hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
  The above diagram is therefore equivalent.\\
  We then have the pullback $E$, such that the following diagram commutes.
  \[
    \begin{tikzcd}
      E \arrow{r}{h} \arrow{d}{e} & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  Since $E$ is a pullback the square above commutes and
  we get $f \circ e = 1_B \circ h = h = 1_B \circ h = g \circ e$, and $E$ is therefore
  indeed an equalizer of $f$ and $g$.
\end{proof}

@@

\begin{theorem}
  \label{arbitrarylimits}
  Let $\mathscr C$ be a category that has all (finite) products and equalizers.
  Let $D : \bm I \to \mathscr C$ be a (finite) diagram.
  Then we can construct a limit of $D$.
\end{theorem}
\begin{proof}
We use the following construction:
\[
  \begin{tikzcd}
    E \arrow{r}{e} &
    \displaystyle \prod_{I \in \bm{I}} D(I)
    \arrow[r,shift left, "s"] \arrow[r,shift right,swap,"t"] &
    \displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D(J)
  \end{tikzcd}
\]
with the maps $s, t$ that have components $s_n, t_n$ for all $K \overset{n}{\to} J \in \bm{I}$:
\begin{align*}
  s_n &= Dn \circ p_K\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_K}{\rightarrow} D_K \overset{Dn}{\rightarrow} D_J \\
  t_n &= p_J\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_J}{\rightarrow} D_J
\end{align*}

The product $\displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D(J)$ has components for each map in the diagram.
We take the diagram for pullbacks as an example for clarification:
\begin{tikzcd}
 I \arrow{r}{m} & J & \arrow[swap]{l}{n} K
\end{tikzcd}.\\
For this example diagram the product has the following projections. One for each identity:
$p_{1_I} \from \prod \to I,\ p_{1_J} \from \prod \to J,\ p_{1_K} \from \prod \to K$,
and one for each remaining map in the diagram: $p_m \from \prod \to J,\ p_n \from \prod \to J$.\\
All products in the above construction exist by assumption.
We take $E$ as the equalizer of $s$ and $t$ which also exists by assumption.
We therefore have $s \circ e = t \circ e$.
The maps $s$ and $t$ have components. The equality has to hold for all
components and we get
\begin{align*}
  s \circ e &= t \circ e\\
  s_n \circ e &= t_n \circ e, \quad \forall\ K \overset{n}{\rightarrow} J \in \bm I \\
  Dn \circ p_K \circ e &= p_J \circ e
\end{align*}
The goal is to construct a limit.
We do this by constructing a cone and showing that it is universal.
For the construction of a cone
we need a vertex and projection maps into every object of the diagram,
such that the commuting condition of cones is satisfied.
We take $E$ as the vertex and
$\uprho_I \coloneqq p_I \circ e, \forall\ I \in \bm I$
as the projection maps of the cone.
Clearly, we have all necessary projection maps and we only need to verify
the commuting condition of cones.
We have all maps $K \overset{n}{\rightarrow} J \in \bm I$
\begin{align*}
  Dn \circ p_K \circ e &= p_J \circ e \\
  Dn \circ \uprho_K &= \uprho_J
\end{align*}
which corresponds to the commuting diagram
\[
  \begin{tikzcd}
    E \arrow{r}{\uprho_K} \arrow{dr}[swap]{\uprho_J} & D_K \arrow{d}{Dn}\\
    & D_J
  \end{tikzcd}
\]
This is exactly the commuting condition we need for a cone.\\
Since any other cone over $D$
is also a fork of $s$ and $t$
it follows that the equalizer $E$ is also a universal cone
and therefore a limit of $D$.
\end{proof}
\begin{remark}
  If $\mathscr C$ has all products including infinite products then
  this construction works even for infinite diagrams.
  If $\mathscr C$ only has finite products, then we can
  use this construction for
  limits of finite diagrams.
\end{remark}

To summarize, we can construct arbitrary limits in the following way.
If $\mathscr C$ has all products and equalizers then we can
construct all limits by
\hyperref[arbitrarylimits]{Theorem \ref*{arbitrarylimits}}.
Equalizer can be constructed using finite products and pullbacks
by \hyperref[arbitrarylimits]{Theorem \ref*{pullbackequalizer}}.
Finite products can in turn be constructed from binary products and a terminal object
by \hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
Binary products can be constructed using a pullback and a terminal object
by \hyperref[arbitrarylimits]{Theorem \ref*{pullbackproduct}}.
Having pullbacks and a terminal object is therefore the lowest requirement from
which it is possible to bootstrap the constructions of other limit instances
and even of arbitrary limits.
